rethinking and the 5 genera

### 2021-03-12 scripts/54_rethinking_binary.Rmd
yeah finally am able to concentrate and make some progress

---
title: "the rethinking way of doing the binary outcome modeling"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

@@

```{r}
meta <- read_csv('../data/amplicon/stool/combined_2_meta.csv') %>% 
  mutate(logdiv_s = scale(log(meta$simpson_reciprocal))) %>% 
  mutate(tox = if_else(toxicity == 'yes', 1, 0),
         cr100 = if_else(cr_d100 == 'yes', 1, 0),
         loca = if_else(center == 'M', 1, 2))
```

```{r}
# visualize the var distribution 
table(meta$cr_d100)  
table(meta$toxicity)
hist(meta$simpson_reciprocal)
hist(log(meta$simpson_reciprocal)) # I think I should use log transformed
table(meta$center)
```
```{r}
dat_list <- list(
    tox = meta$tox,
    cr100 = meta$cr100,
    location = meta$loca,
    logdivs = meta$logdiv_s
    )

mtox <- ulam(
    alist(
        tox ~ dbinom( 1 , p ) ,
        logit(p) <- b*logdivs + a[location] ,
        b ~ dnorm( 0 , 1),
        a[location] ~ dnorm( 0 , 0.5 )
    ) , data=dat_list , chains=4 , log_lik=TRUE )
```

>>>>

---
title: "the rethinking way of doing the binary outcome modeling"
author: "Anqi Dai"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = F, warning = F)
```

@@

# CR/Toxicity ~ alpha diversity + center

## Toxicity

```{r}
meta <- read_csv('../data/amplicon/stool/combined_2_meta.csv') %>% 
  mutate(logdiv_s = scale(log(meta$simpson_reciprocal))) %>%  # diversity is scaled 
  mutate(tox = if_else(toxicity == 'yes', 1, 0),
         cr100 = if_else(cr_d100 == 'yes', 1, 0),
         loca = if_else(center == 'M', 1, 2)) %>%  # MSK 1 ; Upenn 2
  mutate(center = factor(center),
         toxicity = factor(toxicity, levels = c('no','yes')),
         cr_d100 = factor(cr_d100, levels = c('no','yes')))

meta %>% write_csv('../data/amplicon/stool/combined_2_meta_expanded.csv')
```

```{r}
# visualize the var distribution 
table(meta$cr_d100)  
table(meta$toxicity)
#hist(meta$simpson_reciprocal)
#hist(log(meta$simpson_reciprocal)) # I think I should use log transformed
table(meta$center)
table(meta$toxicity, meta$center)
table(meta$cr_d100, meta$center)   
range(meta$logdiv_s)
```

```{r}
set.seed(123)
dat_list <- list(
    tox = meta$tox,
    cr100 = meta$cr100,
    location = meta$loca,
    logdiv_s = meta$logdiv_s
    )

mtox <- ulam(
    alist(
        tox ~ dbinom( 1 , p ) ,
        logit(p) <- b*logdiv_s + a[location] ,
        b ~ dnorm( 0 , 0.25), 
        a[location] ~ dnorm( 0 , 0.5 )
    ) , data=dat_list , chains=4 , log_lik=TRUE , cores = 8)

precis(mtox, depth = 2)
plot(precis(mtox, depth = 2) ) 
# there is baseline difference between penn and Msk and MSK has higher tox rage in general.
# it doesn't seem diversity has a clear association with tox rate after correcting for center wise difference
```

### Prior

```{r}
# doing a prior check
prior <- extract.prior( mtox , n=1e4 ) 
# the diff in the probabilty of yes to toxicity when there are two std diff in diversity in each of the centers 
# for msk  
div_value <- c(-1, 1)
p <- sapply( 1:2 , function(k) inv_logit( prior$b * div_value[k] + prior$a[,1] ) )
mean( abs( p[,1] - p[,2] ) )
# the probability difference is pretty big !  # so reduced the std of the b from 1.5 to 0.25

# for upenn
p <- sapply( 1:2 , function(k) inv_logit( prior$b * div_value[k] + prior$a[,2] ) )
mean( abs( p[,1] - p[,2] ) )
```

### Posterior

```{r}
# posterior prediction check
post <- extract.samples(mtox)
post_tox <- list(m = 1, p = 2) %>% 
  map(function(center_){
    cols = seq(1, length(div_value)) %>% 
      set_names(seq(1, length(div_value))) %>% 
      map(function(idx){
         inv_logit( post$b * div_value[idx] + post$a[, center_] ) 
      })

    ret = mean(abs( cols[[1]] -cols[[2]] ) )
    return(ret)

  })

```

```{r}
# the raw proportion of yes toxicity in each place
rawtox <- meta  %>% 
  count(center, toxicity) %>% 
  left_join(meta %>% 
              count(center, name = 'total'), by = 'center') %>% 
  group_by(center, toxicity) %>% 
  summarise(frac = round(n/total, 2)) %>% 
  filter(toxicity == 'yes')

# don't know how to do the post check since the diversity is continuous variable

```

## CR_d100

```{r}
mcr <- ulam( 
    alist( 
        cr100 ~ dbinom( 1 , p ) ,
        logit(p) <- b*logdiv_s + a[location] ,
        b ~ dnorm( 0 , 1.5), 
        a[location] ~ dnorm( 0 , 0.5 )
    ) , data=dat_list , chains=4 , log_lik=TRUE , cores = 8)

precis(mcr, depth = 2)
plot(precis(mcr, depth = 2) ) 
# there isn't much difference in baseline level between the centers
# diversity seems to be positively associating with CR

# if switching to standardized diversity, the positive association is less sure.

```

```{r}
# doing a prior check
prior <- extract.prior( mcr , n=1e4 ) 
# don't know how to put in logdivs into this prior check
divs_samp <- sample(meta$logdiv_s, size = 1e4, replace = T)
p <- sapply( 1:2 , function(k) inv_logit( prior$b * divs_samp + prior$a[,k] ) )
mean( abs( p[,1] - p[,2] ) )
dens(abs( p[,1] - p[,2] ), adj = 0.1)
```

### 2021-03-12 scripts/55_rethinking_abx.Rmd
yeah finally am able to concentrate and make some progress

meta <- read_csv('../data/amplicon/stool/combined_2_meta_expanded.csv') %>% 
  left_join(m)
```

>>>>

meta <- read_csv('../data/amplicon/stool/combined_2_meta_expanded.csv') %>% 
  left_join(m) %>% 
  mutate(anaerobe_targeting = if_else(is.na(anaerobe_targeting), 'no', anaerobe_targeting)) %>% 
  mutate(anaerobe = if_else(anaerobe_targeting == 'no', 0, 1))

set.seed(123)
dat_list <- list(
    tox = meta$tox,
    cr100 = meta$cr100,
    location = meta$loca,
    anaerobe = meta$anaerobe
    )

meta %>% 
  count(center, anaerobe_targeting) 
```

```{r}   
atox <- ulam(
    alist(
        tox ~ dbinom( 1 , p ) ,
        logit(p) <- b*anaerobe + a[location] ,
        b ~ dnorm( 0 , 1), 
        a[location] ~ dnorm( 0 , 0.5 )
    ) , data=dat_list , chains=4 , log_lik=TRUE , cores = 8)
precis(atox, depth = 2)
plot(precis(atox, depth = 2) ) 
# seems anaerobe abx is positively linked with toxicity 
```

## CR day 100

```{r}
acr <- ulam(  
    alist(
        cr100 ~ dbinom( 1 , p ) ,
        logit(p) <- b*anaerobe + a[location] ,
        b ~ dnorm( 0 , 1), 
        a[location] ~ dnorm( 0 , 0.5 )
    ) , data=dat_list , chains=4 , log_lik=TRUE , cores = 8)
precis(acr, depth = 2)
plot(precis(acr, depth = 2) ) 
# nothing interesting 
```

### 2021-03-12 scripts/56_rethinking_genera.Rmd
yeah finally am able to concentrate and make some progress

set.seed(123)   
dat_list <- list(
    tox = meta$tox,
    cr100 = meta$cr100,
    location = meta$loca,
    A_std = meta$A_std,
    B_std = meta$B_std,
    E_std = meta$E_std,
    F_std = meta$F_std,
    R_std = meta$R_std
    ) 
```

```{r}
gtox <- ulam(
    alist(
        tox ~ dbinom( 1 , p ) ,
        logit(p) <- ba*A_std + bb*B_std + be*E_std + bf*F_std + br*R_std + a[location] ,
        ba ~ dnorm( 0 , 1),
        bb ~ dnorm( 0 , 1),
        be ~ dnorm( 0 , 1),
        bf ~ dnorm( 0 , 1),
        br ~ dnorm( 0 , 1),
        a[location] ~ dnorm( 0 , 0.5 )
    ) , data=dat_list , chains=4 , log_lik=TRUE , cores = 8)
precis(gtox, depth = 2)
plot(precis(gtox, depth = 2) )  
# indicative of the Bacteroides link with toxicity  
```

@@

```{r}
gcr <- ulam(
    alist(
        cr100 ~ dbinom( 1 , p ) ,
        logit(p) <- ba*A_std + bb*B_std + be*E_std + bf*F_std + br*R_std + a[location] ,
        ba ~ dnorm( 0 , 1),
        bb ~ dnorm( 0 , 1),
        be ~ dnorm( 0 , 1),
        bf ~ dnorm( 0 , 1),
        br ~ dnorm( 0 , 1),
        a[location] ~ dnorm( 0 , 0.5 )
    ) , data=dat_list , chains=4 , log_lik=TRUE , cores = 8)
precis(gcr, depth = 2)
plot(precis(gcr, depth = 2) ) 
# I think the Ruminococcus still holds for the CR
```

>>>>

set.seed(123)   
dat_list <- list(
    tox = meta$tox,
    cr100 = meta$cr100,
    location = meta$loca,
    Akkermansia = meta$Akkermansia,
    Bacteroides = meta$Bacteroides,
    Enterococcus = meta$Enterococcus,
    Faecalibacterium = meta$Faecalibacterium,
    Ruminococcus = meta$Ruminococcus

    ) 
```

```{r}
gtox <- ulam(
    alist(
        tox ~ dbinom( 1 , p ) ,
        logit(p) <- ba*Akkermansia + bb*Bacteroides + be*Enterococcus + bf*Faecalibacterium + br*Ruminococcus + a[location] ,
        ba ~ dnorm( 0 , 1),
        bb ~ dnorm( 0 , 1),
        be ~ dnorm( 0 , 1),
        bf ~ dnorm( 0 , 1),
        br ~ dnorm( 0 , 1),
        a[location] ~ dnorm( 0 , 0.5 )
    ) , data=dat_list , chains=4 , log_lik=TRUE , cores = 8)
precis(gtox, depth = 2)
plot(precis(gtox, depth = 2) )  
# indicative of the Bacteroides link with toxicity  
```

@@

```{r}
gcr <- ulam(
    alist(
        cr100 ~ dbinom( 1 , p ) ,
        logit(p) <- ba*Akkermansia + bb*Bacteroides + be*Enterococcus + bf*Faecalibacterium + br*Ruminococcus + a[location] ,
        ba ~ dnorm( 0 , 1),
        bb ~ dnorm( 0 , 1),
        be ~ dnorm( 0 , 1),
        bf ~ dnorm( 0 , 1),
        br ~ dnorm( 0 , 1),
        a[location] ~ dnorm( 0 , 0.5 )
    ) , data=dat_list , chains=4 , log_lik=TRUE , cores = 8)
precis(gcr, depth = 2)
plot(precis(gcr, depth = 2) ) 
# I think the Ruminococcus still holds for the CR
```
